---
title: "assignment3"
author: "ChiaraOrlacchio"
date: "2024-03-09"
output: html_document
---
```{r}
suppressPackageStartupMessages({
  library(AnnotationHub) #to retrived to genome/annotations
  library(Rsubread) #alignment
  library(rtracklayer) #to import/export
  library(Biostrings) #handle sequence
  library(Rfastp) #read trimming & qc
  library(epiwraps) #visualization
  library(GenomicRanges)
})
ah <- AnnotationHub()
```

#We download the data from ENCODE:

```{r, eval=FALSE}
options(timeout=3600) # we need to increase the download timeout since the file is large and cnx slow
dir.create("raw") #new folder where we will put the raw data
download.file("https://www.encodeproject.org/files/ENCFF127RRR/@@download/ENCFF127RRR.fastq.gz", dest="raw/Myc.fastq.gz", mode = "wb") #dest=where to donwload, mode=specifiy binary mode
```


#trimming away adaptor & QC --> to avaoid adaptor contamination
```{r}
dir.create("rfastp.trimmed")
fastq_files <- c(Myc = "raw/Myc.fastq.gz")
qc <- lapply(names(fastq_files), function(x) {
  Rfastp::rfastp(fastq_files[x], thread = 4, overrepresentationAnalysis = TRUE,
                 outputFastq = file.path("rfastp.trimmed/", gsub("\\.fastq\\.gz$", "", basename(fastq_files[x]))))
})
#x-> contains path to each of the fastq files -> makes it easier when I have a lot of files to process
```

# Alignment

## Using Rsubread

### Building a genome index for mapping

This we have to do only once for a genome, and can then re-use across projects

```{r, eval=FALSE}
# we get the genome sequence from AnnotationHub
drosophila_genome <- ah[["AH49674"]]
# we create a new directory that will contain the genome index
dir.create("BDGP6_genome")
# we write the genome sequence in fasta format
export(import.2bit(drosophila_genome), "BDGP6_genome/genome.fasta.gz", compress=TRUE)
# we build a Rsubread index
Rsubread::buildindex("BDGP6_genome/rsubread", reference="BDGP6_genome/genome.fasta.gz") #"location", “genome.file"
```

### Alignment

```{r}
dir.create("aligned")
align.stats <- Rsubread::align(index="BDGP6_genome/rsubread", type="dna",
                               readfile1=c("rfastp.trimmed/Myc_R1.fastq.gz"),
                               output_file=c("aligned/Myc.bam"),
                               nthreads=6, sortReadsByCoordinates=TRUE) #bam files conatin alignment
align.stats 
list.files("aligned")
```

#### Calculate the number and percentage of mapped reads
```{r}
mapped_reads <- align.stats$Myc.bam[2]
total_reads <- align.stats$Myc.bam[1]
mapping_percentage <- (mapped_reads / total_reads)*100

cat("Mapped Reads:", mapped_reads, "\n")
cat("Mapping Percentage:", mapping_percentage, "%\n")
```
## How many peaks were found and plot the signal around one of the peaks that is located inside a gene
#drosophila gene loading in ensdb format
```{r}
q <- query(ah, c("Drosophila", "EnsDb"))
q
drosophila_ensdb <- ah[["AH116255"]]
genes <- transcriptsBy(drosophila_ensdb,"gene")
x <- GRanges(unlist(genes)) #to make get rid of the list format of the "genes"
```

```{r}
dir.create("peaks")
peaks <- callPeaks("aligned/Myc.bam", fragLength=50L)

# if we want to save it as a bed file:
rtracklayer::export.bed(peaks, "peaks/peaks.bed")

# Number of peaks found
num_peaks <- length(peaks)
cat("Number of Peaks Found:", num_peaks, "\n")

# Get the indices of peaks that overlap with genes
peaks_overlap <- peaks[overlapsAny(peaks,x)]
peaks_overlap

#plot peak inside gene
plotSignalTracks(c(Myc="aligned/Myc.bam"), region=peaks_overlap[1],
                 extend=1000,ensdb = drosophila_ensdb,genomeAxis = 1,)





```
