---
title: "assignment5"
author: "ChiaraOrlacchio"
date: "2024-03-24"
output: html_document
---
```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(rtracklayer)
  library(epiwraps)
  library(GenomicRanges)
})
ah <- AnnotationHub()
```

#downloading the histone peaks
```{r}
dir.create("peaks")

# H3K4me3
download.file("https://www.encodeproject.org/files/ENCFF671UNN/@@download/ENCFF671UNN.bed.gz", dest="peaks/H3K4me3.bed.gz")

peaks_H3K4me3 <- rtracklayer::import("peaks/H3K4me3.bed.gz", format = "narrowPeak")

# H3K27me3
download.file("https://www.encodeproject.org/files/ENCFF055QNY/@@download/ENCFF055QNY.bed.gz", dest="peaks/H3K27me3.bed.gz")

peaks_H3K27me3 <- rtracklayer::import("peaks/H3K27me3.bed.gz", format = "narrowPeak")
```

#Using the peaks you downloaded last week, identify bivalent domains (H3K27me3 + H3K4me3) in mouse embryonic stem cells (mESC)

```{r}
bivalent_overlaps <- findOverlaps(peaks_H3K4me3, peaks_H3K27me3, type = "any")
bivalent_domains <- peaks_H3K4me3[queryHits(bivalent_overlaps)] #Diese Funktion verwendet die Ergebnisse der findOverlaps-Funktion, um die tatsächlichen Bereiche (Peaks) aus dem H3K4me3-Set zu extrahieren, die mit den identifizierten bivalenten Domänen überlappen.

cat("There are", length(bivalent_domains), "overlaps of H3K4me3 with H3K27me3")
```
##Question: what happens to those regions upon differentiation?

```{r}
#Choose one differentiated cell type (e.g. hepatocytes, neural progenitor, or smooth muscle      cells)
#Download the H3K27me3 and H3K4me3 peaks from this cell type

# H3K4me3_hep
download.file("https://www.encodeproject.org/files/ENCFF252RRE/@@download/ENCFF252RRE.bed.gz", dest="peaks/H3K4me3_hep.bed.gz")

peaks_H3K4me3_hep <- rtracklayer::import("peaks/H3K4me3_hep.bed.gz", format = "narrowPeak")

# H3K27me3_hep
download.file("https://www.encodeproject.org/files/ENCFF290NCY/@@download/ENCFF290NCY.bed.gz", dest="peaks/H3K27me3_hep.bed.gz")

peaks_H3K27me3_hep <- rtracklayer::import("peaks/H3K27me3_hep.bed.gz", format = "narrowPeak")
```

#How many of the mESC bivalent domains are, in this differentiated cell type, overlapping either mark or their combination (in this differentiated cell type)?
```{r}
# H3K4me3_hep
bi_overlaps_h3k4me3_hep <- findOverlaps(bivalent_domains, peaks_H3K4me3_hep) #identifies overlaps between the H3K4me3 peaks in the differentiated cell type (h3k4me3_hep_gr) and the bivalent domains identified in mESCs (bivalent_domains)

bi_do_h3k4me3_hep <- bivalent_domains[queryHits(bi_overlaps_h3k4me3_hep)] #exact locations (peaks) where overlaps occur

cat("After differentiation into hepatocytes", length(unique(bi_do_h3k4me3_hep)), "bivalent domains overlap with H3K4me3") #number of bivalent domains overlapping with H3K4me3 in the differentiated cell type
```

```{r}
# H3K27me3_hep
bi_overlaps_h3k27me3_hep <- findOverlaps(bivalent_domains, peaks_H3K27me3_hep)

bi_do_h3k27me3_hep <- bivalent_domains[queryHits(bi_overlaps_h3k27me3_hep)]

cat("After differentiation into hepatocytes", length(unique(bi_do_h3k27me3_hep)), "bivalent domains overlap with H3K27me3") #number of bivalent domains overlapping with H3K27me3 in the differentiated cell type
```

#or their combination (in this differentiated cell type)?
```{r}
bi_overlap_hep <- findOverlaps(peaks_H3K4me3_hep, peaks_H3K27me3_hep)
bi_domain_hep <- peaks_H3K4me3_hep[queryHits(bi_overlap_hep)]
bi_overlaps_combi <- findOverlaps(bivalent_domains, bi_domain_hep)
bi_domains_combi <- bivalent_domains[queryHits(bi_overlaps_combi)]

cat("In the differentiated cell type, there are", length(unique(bi_domains_combi)), "bivalent domains where both H3K4me3 and H3K27me3 overlap") #number of bivalent domains where both H3K4me3 and H3K27me3 overlap in the differentiated cell type
```

```{r}
#Interpretation: After differentiation we have generally less bivalent regions since the function of the cells are no more "clear" and the need to be ambivalent gets less and less (one cell type has specific functions). It can also be observed that there are more overlaps between the bivalent domains with the H3K27me3 (repressive properties) than with the H3K4me3.
```


